---
title: "Data Cleaning Cyclistic 2016 Trips"
author: "Fábio Lamaison Pinto"
date: "2024-01-01"
output:
  html_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load Library
library(tidyverse)
library(janitor)
library(ggplot2)
library(here)
library(lubridate)

# Functions

is_only_customer <- function(null_values) {
  if(null_values[1,1] == "Customer") {
    print("True")
  } else {
    print("False")
  }
}
```

## DATA CLEANING AND VALIDATION

###### [FOR THE GOOGLE PROFESSIONAL DATA ANALYTICS CERTIFICATE CAPSTONE PROJECT - TRACK 1]{.smallcaps}

This is a capstone project for the Google Data Analytics Professional
Certificate it'll be based on a fictional bike sharing company called
Cyclistic, which uses Divvy's data sets on the use of their services as
'proxy' data <https://divvy-tripdata.s3.amazonaws.com/index.html>.

The tool chosen to do the analysis is the R markdown, as you can see,
this tool was chosen due to it's capabilities in clearly showing the
whole process, and allowing peers to verify step by step the process
taken to reach the end result, SQL and Google Sheets where also explored
as a option, and part of the analytics process where developed on them,
but further discarded as R presented to be a sturdier choice, providing
an all in one solution.

The data provided for the year 2016, comprehends .csv files of tables on
trips and stations, for this project we'll focus only on trips data
frames, they are divided on quarters, being the second quarter divided
on it's months, so first of all, we are going to load all of them and
join them on a single data frame.

```{r}
trips_Q1 <- read_csv("~/Data/Trips/source/Divvy_Trips_2016_Q1.csv")

trips_2016_04 <- read_csv("~/Data/Trips/source/Divvy_Trips_2016_Q1Q2/Divvy_Trips_2016_04.csv")
trips_2016_05 <- read_csv("~/Data/Trips/source/Divvy_Trips_2016_Q1Q2/Divvy_Trips_2016_05.csv")
trips_2016_06 <- read_csv("~/Data/Trips/source/Divvy_Trips_2016_Q1Q2/Divvy_Trips_2016_06.csv")

trips_Q2 <- rbind(trips_2016_04, trips_2016_05, trips_2016_06)

trips_Q3 <- read_csv("~/Data/Trips/source/Divvy_Trips_2016_Q3.csv")

trips_Q4 <- read_csv("~/Data/Trips/source/Divvy_Trips_2016_Q4.csv")

trips_2016 <- rbind(trips_Q1, trips_Q2, trips_Q3, trips_Q4)
```

Now that all of them are joined, it's time to verify if the data is
cleaned and ready to analyze, for the first step on cleaning, metadata
and a summarized description of the data-set will be presented, and
serve as a base to guide the cleaning process, metadata on Divvy
data-set on trips is as follows:

> Variables:

> trip_id: ID attached to each trip taken starttime: day and time trip
> started, in CST stoptime: day and time trip ended, in CST bikeid: ID
> attached to each bike tripduration: time of trip in seconds
> from_station_name: name of station where trip originated
> to_station_name: name of station where trip terminated
> from_station_id: ID of station where trip originated to_station_id: ID
> of station where trip terminated usertype: "Customer" is a rider who
> purchased a 24-Hour Pass; "Subscriber" is a rider who purchased an
> Annual Membership gender: gender of rider birthyear: birth year of
> rider

> Notes:
>
> -   First row contains column names
>
> -   Trips that did not include a start or end date were removed from
>     original table.
>
> -   Gender and birthday are only available for Subscribers

Trips Table Data-set Summary.

```{r}
summary(trips_2016)
```

The summary shows that the table data-set for trips has 13 columns, with
entries relative to trips on IDs, date time, stations, and user info.

### On a first glance, is possible to verify:

-   All the 12 columns described on meta data are present in the data
    frame.
-   Column names are semantically appropriated, but don't follow the
    same name convention.
-   Time related columns are stored as character type, optimally they
    should be DateTime types.
-   A bike ID column is present, but there is no bike data-set to be
    consulted.
-   There is no user ID, nor an user data set, what limits the capacity
    to spot incorrect entries for user related fields.

### Ensuring Parameters Depicted on Metadata:

Starting from this, it's important to further analyse the data set,
ensuring to spot potentially hidden inconsistencies, that may lead to
wrong outcomes, having this in mind, I'll dive deeper and check for null
values, but first, lets ensure that all column names follow the snake
case convention.

```{r}
colnames(trips_2016) <- c("trip_id","start_time","stop_time","bike_id","trip_duration","from_station_id" ,"from_station_name","to_station_id","to_station_name","user_type","gender","birth_year")
colnames(trips_2016)
```

Snake case ensured to all the column names, time to check for null
values.

```{r}
print(colSums(is.na(trips_2016)))
```

As was informed in meta data, null values can be present for gender and
birthday, only if they belong to customer user type entries, so let's
make sure nulls are only present on customer's entries.

```{r}
null_gender <- trips_2016 %>%
  filter(is.na(gender)) %>%
  group_by(user_type) %>%
  summarize(count = n())

print("Are null values for gender only present for the Customer User Type?")
is_only_customer(null_gender)

null_birth_year <- trips_2016 %>%
  filter(is.na(birth_year)) %>%
  group_by(user_type) %>%
  summarize(count = n())

print("Are null values for birtheyear only present for the Customer User Type?")
is_only_customer(null_birth_year)
```

Checking if there are non null values for Customers birth year and
gender.

```{r}
# Check for the whole year
trips_2016 %>%
  filter(user_type == "Customer") %>%
  select(user_type, birth_year, gender) %>%
  na.omit() %>%
  head()
```

Some customers have non null values, even thought metadata on the
data-set says that they shouldn't, there are multiple hypothesis on why,
but since there isn't a way to validate them with the ones responsible
for the data, this entries will be discarded.

```{r}
trips_2016 <- trips_2016 %>%
  filter(!(user_type == "Customer" & (!is.na(gender) | !is.na(birth_year))))

summary(trips_2016)
```

### Birth Year Entries:

Reviewing the summary, There is some curious values for birth year, that
endure even after dropping what should be the wrong entries for birth
year on the data set, these curious values are such as an informed birth
year of 1899 (117 years) to deal with this I'll start by checking the
whole range of values for birth year.

```{r}
birth_year_range <- range(na.omit(trips_2016$birth_year))
print(birth_year_range)
```

The range goes from 1899 to 2000, taking in consideration that the
data-set I'm working with has data relative to 2016, this means that at
that time users would be of ages spanning from.

```{r}
age_range <- 2016 - birth_year_range 
print(age_range)
```

When it comes to the youngest age, 16 seems fine, as there was no limit
of age described for being a subscribers or customer from the service,
but when it comes to the oldest, 117 years of age seems too far, even
for a user that opts for assistive options, so lets check out how often
this users appear, and explore ways to check if this is a valid entry,
or an error.

```{r}
cyclistic_supercentenarians <- trips_2016 %>%
  filter(birth_year < (1907)) %>%
  transform(age = 2016 - birth_year) %>%
  select(-birth_year)

print("Ages and their entries")
print(table(cyclistic_supercentenarians$age))
print(paste("Total entries",length(cyclistic_supercentenarians$birth_year)))
```

#### Supercentenarians:

There are 442 entries of ages equal or above 110 years, that can be
grouped on 4 different values, being them 110, 115, 116, and 117, with a
116 being the most common entry, with 203 occurrences, it is important
to notice that entries do not directly correspond to amount of users,
since multiple entries can correspond to a single user, but different
ages mean that at least one user of each must exist, once again the
question surges, are these inputs reliable? could people of such age
have the physical capacity to use the bike sharing service, even when it
comes to assistive options?

One step to check this would be to query the ID of assertive bike
models, and compare them to the bike IDs present on the entries,
unfortunately access to bike ids data-set isn't provided by this project
fictional company Cyclistic, but since the company declares that 8% of
it's users use assistive options, and supercentenarian users (people of
110 years or more) account for less than 1% on the trip entries, they
could be part of this users percentage, and these entries should be
leaved it as it is, right?

#### Gerontology Research Group:

Turns out that there is a research team dedicated to keep track of
supercentenarians around the world, they are the Gerontology Research
Group <https://grg.org/>, and they have a data-set on supercentenarians
<https://grg.org/WSRL/TableE.aspx>, that will be brought to this
analysis as a way to validate the existence of such entries, and point
if birth year entries should be further verified by the ones in charge
of the data, that presumably have more resources to further verify the
existence of wrong entries in other age ranges.

##### Supercentenarians data-set structure:

The hosted data-set has 3 different tables, one for supercentenarians
proven to be alive on their last research, one for proven the be
deceased on their last research, and one for "Limbo" supercentenarians
that were once registered and verified, but that they lost contact with
as their last research, having their status unknown, below the table's
column names are shown to provide a glimpse oh their structure.

```{r}
alive_sc_2023 <- read.csv("~/Data/Supercentenarians/alive_supercentenarians_2023.csv")

deceased_sc_2023 <- read.csv("~/Data/Supercentenarians/deceased_supercentenarians_2023.csv") 

unknown_sc_2023 <- read.csv("~/Data/Supercentenarians/unknown_supercentenarians_2023.csv") 

print("Alive on december 11 2023")
colnames(alive_sc_2023)
print("Deceased before december 11 2023")
colnames(deceased_sc_2023)
print("Unknown state on december 11 2023")
colnames(unknown_sc_2023)
```

As it shows, the tables have different columns, for the alive table
there is the rank column, for the deceased table there is Deathplace and
Died columns, and for the Unknown table there isn't a Rank, Deathplace
or Died column, we will further investigate this data sets structure,
but for now they will be joined into one data-set to ease the process of
cleaning and transforming them.

##### Supercentenarians data cleaning and transforming:

The tables will be joined keeping only columns that have variables that
can be related to our trips data-set, such as age, sex (gender),
residence or Deathplace, died (death year), and application date or
added date to back the eventual lack of entry for application date (for
unknown state supercentenarians, this is a date were they were certainly
alive); a column called alive will be added to the data-sets to store
information that was explicit on their titles, but wasn't on the tables,
and that will serve to store a Boolean validation on which of the
subjects were alive in 2016.

```{r}
alive_sc_2023 <- alive_sc_2023 %>%
  select(Residence, Age, Sex, Applic.Date, Date.Added) %>%
  mutate(Alive = TRUE)

deceased_sc_2023 <- deceased_sc_2023 %>%
  transform(Date.Added = DateAdded) %>%
  select(Deathplace, Died, Age, Sex, Applic.Date, Date.Added)

unknown_sc_2023 <- unknown_sc_2023 %>%
  select(Residence, Age, Sex, Applic.Date, Date.Added)

grg_supercentenarians <- bind_rows(alive_sc_2023, deceased_sc_2023, unknown_sc_2023)

summary(grg_supercentenarians)
```

Lets clean and transform column names, so they are more consistent and
follow the same pattern (snake case) also lets join residence and died
(death place) columns in one column called last_location, so it will be
easier to identify if their last known location is close or in Chicago,
what would strengthen the possibility of them being subscribers of the
Cyclistic bike sharing service.

```{r}
grg_supercentenarians <- clean_names(grg_supercentenarians)

grg_supercentenarians <- grg_supercentenarians %>%
  transform(deceased_date = died) %>%
  transform(last_location = ifelse(is.na(deathplace), residence, deathplace)) %>%
  transform(applic_date = ifelse(is.na(applic_date), date_added, applic_date)) %>%
  select(-deathplace, -died, -residence, -date_added)

head(grg_supercentenarians)
```

As the data frame above shows, the data is now structured as a data
frame with 6 columns, coming from tables that had a maximum of 12
columns, this selection allows to focus on the most important values for
the verification processes that will follow up, and to save time and
computing resources, next I'll transform date stored values from
character to lubridate date time format, and also isolate the year from
the age value.

```{r}
grg_supercentenarians <- grg_supercentenarians %>%
  transform(applic_date = mdy(applic_date)) %>%
  transform(deceased_date = mdy(deceased_date)) %>%
  transform(age = sub(" .*", "", age))

head(grg_supercentenarians)
```

#### Paired Analysis.

Having the data cleaned and ready to be used to verify the validity of
supercentenarians related entries on the trips data-set, is time to
start the process, first I'll filter last known location to include only
U.S.A. entries, unfortunately, as there isn't a standard format for
inputs related to the United States, I will try to narrow down entries
by filtering for locations that start with the letter "U".

```{r}
grg_supercentenarians <- grg_supercentenarians %>%
  filter(startsWith(last_location, "U")) %>%
  print()
```

The effort was succesful by lowering the total entries to 192, and
including other than the desired entry, values related to the United
Kingdom, as a next step I will filter out the "UK" entries for location
as also filter out the deceased before 2016, in the case of lack of an
entry for deceased date, I'll use application date instead.

```{r}
grg_supercentenarians <- grg_supercentenarians %>%
  filter(!startsWith(last_location, "UK")) %>% 
  filter(alive == TRUE | ifelse(is.na(deceased_date),year(applic_date) > year(ym("2015-01")), year(deceased_date) > year(ym("2015-01")))) %>%
  print()
```

Dropped more 125 values, lets see if it is possible to also filter on
gender, for it to be possible entries for gender on the Cyclistic trips
data frame would need to entirely present only one gender value for
every row, I'll do this by retrieving the values for gender in the
Cyclistic supercentenarians data frame, formatted as simply "M" or "F"
to be compatible with the values for sex in the GRG supercentenarians
data frame, and them simply check the existence of matching values
between the two.

```{r}
cyclistic_sc_gender <- cyclistic_supercentenarians %>%
  distinct(gender) %>%
  mutate(gender = substr(gender,1,1))

grg_supercentenarians <- grg_supercentenarians %>%
  filter(sex %in% cyclistic_sc_gender) %>%
  print()
```

Strangely, there are indeed only entries for male supercentenarians on
the Cyclistic trips data frame, once females are filtered out of the GRG
supercentenarians data frame, there are only 2 subjects left, with only
one of them having lived in Illinois - IL , the state where Chicago is
located, the other lived in the other side of the country, lets end the
transformations by calculating how old they would be in 2016.

```{r}
grg_supercentenarians %>%
  transform(age = (as.integer(age) - (year(deceased_date) - 2016))) %>%
  print()
```

The calculation resulted in the only possible male supercentenarian
living close or in Chicago to not be a supercentenarian (110 or older)
in 2016, backing up the hypothesis that the entries for
supercentenarians aren't valid entries, and in fact are wrong inputs,
what points to the possibility that more entries on birth year can be
unreliable, and should be further investigated by data owners.

On what is possible to be sure and corrected around birth year entries
on the trips data-set, every entry depicting users of 110 years of age
or older will be dropped.

```{r}
trips_2016 <- trips_2016 %>%
  filter(birth_year > 1906 | is.na(birth_year))

print(trips_2016)

```

#### Filtering Undesired Data.

Going back to focus only on the Cyclistic trips data frame, and
recalling that it's not a possibility to consult a bike data-set through
bike ID, it's clear that the use of the bike id column is greatly
limited, due to this it will be dropped to help keep in the data frame
only information that will be valuable for the analysis step.

```{r}
trips_2016 <- trips_2016 %>%
  select(-bike_id)

head(trips_2016)
```

Checking for duplicates on trip ID.

```{r}
trip_dupes <- trips_2016 %>%
  get_dupes(trip_id)

print(trip_dupes)
```

Returned 50 duplicated entries, I'll drop them to keep only distinct
entries.

```{r}
trips_2016 = trips_2016 %>%
  distinct(trip_id, .keep_all = TRUE)

```

Next I will check for rides with a length smaller or equal to 30
seconds, as a measure to square out possible situations where the bikes
were taken from their station and shortly after given back, without
properly resulting in a trip, and even possible errors in time stamps
that would end up depicting trips that took place between different
stations but have entries with trip lengths that are impossible to
achieve.

```{r}
trips_2016 %>%
  filter(trip_duration < 31) %>%
  print()
```

None values smaller than 31 seconds were found, resulting in no need to
further cleaning related to this subject, as far as it is possible to
evaluate with the given data and R libraries, all entries seeem
consistent when it comes to trip length.

#### Transforming Desired Columns.

Lets rearrange date time related columns, so the info is presented as
start date (Date type), start time (time type), and duration (time type)
starting on the start time column, it will be verified for entry
formats.

```{r}
trips_2016 %>%
  mutate(start_time_length = nchar(start_time)) %>%
  group_by(start_time_length) %>%
  slice(1) %>%
  ungroup() %>%
  print(trips_2016$start_time)

```

There is two different formats, both of them follow the mm/dd/YYYY
format, but some have seconds in their time stamp, while others don't,
lets add a string with ":00" as seconds, to the ones that don't follow
the HH:MM:SS standard.

```{r}
trips_2016 <- trips_2016 %>%
  mutate(start_time = ifelse(nchar(start_time) < 17, paste(start_time, ":00", sep=""), start_time))

trips_2016 <- trips_2016 %>%
  mutate(stop_time = ifelse(nchar(stop_time) < 17, paste(stop_time, ":00", sep=""), stop_time))

trips_2016 %>%
  group_by(nchar(start_time)) %>%
  slice(1) %>%
  ungroup() %>%
  print(trips_2016$start_time)

```

Done, now its time to turn these strings in date time format, to after
calculate the time difference between them.

```{r}
trips_2016 <- trips_2016 %>%
  mutate(start_time = mdy_hms(start_time)) %>%
  mutate(start_time = ymd_hms(start_time)) %>%
  mutate(stop_time = mdy_hms(stop_time)) %>%
  mutate(stop_time = ymd_hms(stop_time)) %>%

print(trips_2016)
```

The execution was successful, but with some warnings, lets check for the
values.

```{r}
trips_2016 %>%
  group_by(nchar(start_time)) %>%
  slice(1) %>%
  ungroup() %>%
  print(trips_2016$start_time)

trips_2016 <- trips_2016 %>%
  filter(!is.na(start_time))

trips_2016 %>%
  group_by(nchar(start_time)) %>%
  slice(1) %>%
  ungroup() %>%
  print(trips_2016$start_time)
```

All done, null values where present, but now they are filtered out, I'll
follow the process by calculating and storing the trip duration, as it
will be useful on the analysis step to calculate metrics related to the
trip duration.

```{r}
trips_2016 <- trips_2016 %>%
  mutate(trip_duration = stop_time-start_time)

head(trips_2016)
```

To finish the data cleaning of the trips data-set, time entries will be
restructured on 3 columns, as previously discussed, also the birth year
column will pass through transformation to store user age instead of
birth year to facilitate understanding, it will be named appropriately.

```{r}
trips_2016$start_date <- date(trips_2016$start_time)
trips_2016$start_time <- format(trips_2016$start_time, format="%H:%M:%S")
trips_2016$age <- (2016-trips_2016$birth_year)

trips_2016 <- trips_2016 %>%
  select(trip_id, start_date, start_time, trip_duration,from_station_name, from_station_id, to_station_name, to_station_id, user_type, gender, age) 
head(trips_2016)
```

#### Manual Verification.

For peace of mind, lets check alphabetically arranged distinct values
for stations, to make sure that there aren't some misspells on their
names that could lead to entries for the same station not being
grouped/summarized when needed.

```{r}
trips_2016 %>%
  distinct(from_station_name, .keep_all = TRUE) %>%
  arrange(from_station_name) %>%
  print()
```

```{r}
trips_2016 %>%
  distinct(to_station_name, .keep_all = TRUE) %>%
  arrange(to_station_name) %>%
  print()
```

After verifying every distinct entry, no mistakenly inserted value for
station names was found, a curious property for station name was
noticed, "(...)" is present after station names, apparently it is used
to make notations on stations, some only contain a "\*" without further
information to be found in any of the data provided, while others have
more detailed information as "(NEXT Apts)" which upon online search
point out to be a building that serves as a point of reference.

#### Stations Data-set.

Additionally I'll bring the data-sets on stations for it to be cleaned
and include what may be of best use for the analysis step, the data-sets
on stations are also divided in 3 different data-sets, one for Q1 and
Q2, and the other two for Q3 and Q4 respectively, they will be joined in
one, storing only unique values, and the columns of interest will be the
ones related to name, id, and coordinates.

```{r}
stations_Q1Q2 <- read_csv("~/Data/Stations/source/Divvy_Stations_2016_Q1Q2.csv")
stations_Q3 <- read_csv("~/Data/Stations/source/Divvy_Stations_2016_Q3.csv")
stations_Q4 <- read_csv("~/Data/Stations/source/Divvy_Stations_2016_Q4.csv")
stations <- rbind(stations_Q1Q2, stations_Q3, stations_Q4)
stations <- unique(stations)

print(stations)
```

Two from the six columns can be discarded, as they refer to the station
capacity and the day they went online, none of these will be of good use
for what I'm proposing to analyse.

```{r}
stations_coord <- stations %>%
  select(-online_date, -dpcapacity)
print(stations_coord)
```

### Conclusion.

That's it, the cleaning and basic transformation processes are done,
data was verified, what was spotted as bad inputs based on metadata were
discarded, column names were put to a standard format (snake case) based
on the existing most common one, birth year was transformed to age as it
is an easier format to present, centenarians entries for age were put to
test using another data-set what led then to be proven invalid and then
discarded, duplicates were cleaned, and possible inconsistencies that
could be further cleaned upon consulting data owners were stated.
